\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-11 pldi/draft.tex
Pass on 5.1

To address this challenge, we develop \sys, a shell parallelizing\todo{/that parallelizes} POSIX shell scripts.
\kk{Mention that it outputs shell scripts.} % to their parallel equivalents.
% \sys offers both classes of users in the second group such much-needed automation.
\sys benefits both programmer groups mentioned earlier, with a particular emphasis on users.
Command developers are provided a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
  rather than providing a command's full observable behavior, these annotations focus primarily on their interaction with state.
Shell users, on the other hand, are provided full automation:
  \sys extracts parallelism latent in their scripts by analyzing their parallelizability properties. % of their fragments.
% benefiting scripting experts and non-experts alike by offering automated POSIX-aware parallelization of shell-scripts as long as command developers
\sys's transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will choose to not improve performance rather than risking breakage.

@@

@@

As mentioned in \Cref{ir}, in order to optimize a shell script, \sys
translates it to a dataflow graph.  However, there are several
components in shell programs that cannot be arbitrarily parallelized
without affecting the program behaviour---\eg commands connected with
a \ttt{\&\&} operator~\sx{bg}. In addition, the shell's highly dynamic
nature compounds the challenge---\eg by not knowing what files does an
unexpanded command argument refer to.
% of correctly identifying what parts can be distributed, e.g.

@@

To address the first issue, we introduce the notion of parallelizable
program regions and a translation pass for converting them to a
dataflow graph. To address the second issue, we limit dynamic
components by deferring translation as late as possible using a
wrapping scheme.

@@

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  cat f1 f2 | grep ``foo'' > f3 && sort f3 
\end{lstlisting}
% # Replace sort with sth better

\noindent
The commands \ttt{cat f1 f2} and \ttt{grep ``foo''} would execute as
independent processes in the standard shell, while \ttt{sort f3} would
wait for their completion before executing.  Both \ttt{cat f1 f2 |
  grep ``foo'' > f3} and \ttt{sort f3} are therefore parallelizable
regions, regions that cannot be extended beyond \ttt{\&\&}---even
better, they are maximal.

Intuitively, parallelizable regions correspond to sub-expressions of
the program that would be allowed to execute independently by
different processes in the POSIX standard~\cite{posix}. Larger
parallelizable regions can be composed from smaller ones using the
pipe operator \ttt{|} and the parallel composition operator \ttt{\&}.
Conversely, the \ttt{;} sequential composition operator, the
\ttt{\&\&} logical operator, and function definitions represent
barrier constructs that do not allow parallelizable regions to
permeate through.

@@

\kk{It might be a good idea to change this subsection based on the new
  as-late-as-possible architecture where everything is wrapped and
  translated at the final moment. Ideally, this would become the main
  point of this subsection, as the translation algorithm is kind of
  redundant when having the distributable regions too.}

The \sys front-end performs a depth-first search pass on the AST of
the given shell program.  During this pass, it greedily extends the
distributable regions from the bottom up, translating their
independent components to dataflow graph nodes until a barrier
construct is reached.  To identify possible opportunities for
parallelism, the translation pass also extracts the category of each
node together with its input and output pairs.  To perform this
extraction, it uses the command name to index into a key-value map
with values that are expressed in the DSL for distributability
characterization~\sx{ext}.

A few illustrative cases of the analysis algorithm are shown in pseudocode below:

\begin{lstlisting}[language=python, float=h]
  def translate(node):
    ...
    elif(node.name == 'Pipe'):
      return pipe_graphs([translate(child)
        for child in node.children])
    elif(node.name == 'And'):
      node.children = [translate(child)
        for child in node.children]
      return node
    elif(node.name == 'Command'):
      if(not safe(node)):
        return make_ir_node(node, 'unsafe')
      else:
        return make_ir_node(node,
          find_category(node))
    ...
\end{lstlisting}

Whenever the algorithm encounters a distributable node, it recursively
translates its subcomponents.
It then merges them in a dataflow (sub-)graph, connecting the output of the first with the input of the second
% what first and second? we didn't say anything about first and second
Before the different dataflow graphs are connected, the
algorithm checks that at most one node in the graph writes at every
file, and that at most one node reads from a file that another node writes
to. This is important, in order to avoid inconsistent behaviour after
distributing nodes in the dataflow graph (\ie by introducing to
concurrent reads and writes to the same file when
parallelizing).

Whenever the algorithm encounters a barrier node,
it recursively translates its subcomponents, similarly to the distributable
case.
However, different from the distributable case, it does not merge them in a large dataflow graph.

Finally, whenever the algorithm encounters a command, it first
checks whether it is safe to parallelize it. A command is considered
safe for parallelization if (i) it belongs to the \sta or \pur distributability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this. % why couldn't we replicate this?
If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.

An important consideration is that due to the highly dynamic nature of
the shell, the safety properties that are mentioned above cannot be
soundly checked statically.  For safety purposes, \sys takes a
conservative approach where it does not parallelize nodes without
complete information.  For example, the translation pass cannot infer
that two un-expand strings do not refer to the same file identifier,
and as a result does not parallelize this sub-expression.

@@

\kk{Maybe guide all of this by showing an example of an output shell
  script?}

@@

>>>>

To address this challenge, we develop \sys, a shell parallelizing\todo{/that parallelizes} POSIX shell scripts.
\kk{Mention that it outputs shell scripts.} % to their parallel equivalents.
% \sys offers both classes of users in the second group such much-needed automation.
\sys benefits both programmer groups mentioned earlier, with a particular emphasis on users.
Command developers are provided a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
  rather than providing a command's full observable behavior, these annotations focus primarily on their interaction with state.
Shell users, on the other hand, are provided full automation:
  \sys extracts parallelism latent in their scripts by analyzing their parallelizability properties. % of their fragments.
% benefiting scripting experts and non-experts alike by offering automated POSIX-aware parallelization of shell-scripts as long as command developers
\sys's transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will choose to not improve performance rather than risking breakage. \kk{Should we mention that we only show this empirically?}

@@

\kk{Make sure for the running script to contain more than just a
  pipeline, so that it is clear that \sys turns a script into the same
  script with DFGs in place of ``parallelizable-regions'' (which are
  usually pipelines).}

@@

As mentioned in \Cref{ir}, in order to optimize a shell script, \sys
translates it to a dataflow graph.  However, there are several
components in shell programs that cannot be arbitrarily parallelized
without affecting the program behaviour---\eg commands connected with
a \ttt{\&\&} operator~\sx{bg}.
%% In addition, the shell's highly dynamic
%% nature compounds the challenge---\eg by not knowing what files does an
%% unexpanded command argument refer to.
To address this issue, we introduce the notion of parallelizable
program regions and a translation pass for converting them to dataflow
dataflow graphs.

@@

%% To address the first issue, we introduce the notion of parallelizable
%% program regions and a translation pass for converting them to a
%% dataflow graph. To address the second issue, we limit dynamic
%% components by deferring translation as late as possible using a
%% wrapping scheme.

@@

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  cat f1 f2 | grep "foo" > f3 && sort f3 
\end{lstlisting}
% # Replace sort with sth better

\noindent
The commands \ttt{cat f1 f2} and \ttt{grep "foo"} would execute as
independent processes in the standard shell, while \ttt{sort f3} would
wait for their completion before executing.  Both \ttt{cat f1 f2 |
  grep "foo" > f3} and \ttt{sort f3} are therefore parallelizable
regions, regions that cannot be extended beyond \ttt{\&\&}---even
better, they are maximal.

Intuitively, parallelizable regions correspond to sub-expressions of
the program that would be allowed to execute independently by
different processes in the POSIX standard~\cite{posix}. Larger
parallelizable regions can be composed from smaller ones using the
pipe operator \ttt{|} and the parallel composition operator \ttt{\&}.
Conversely, all other operators, such as the \ttt{;} sequential
composition operator, the \ttt{\&\&} logical operator represent
barrier constructs that do not allow parallelizable regions to
permeate through.

@@

%% \kk{It might be a good idea to change this subsection based on the new
%%   as-late-as-possible architecture where everything is wrapped and
%%   translated at the final moment. Ideally, this would become the main
%%   point of this subsection, as the translation algorithm is kind of
%%   redundant when having the distributable regions too.}

The \sys front-end performs a depth-first search pass on the AST of
the given shell program.  During this pass, it extends the
parallelizable regions from the bottom up, translating their
independent components to dataflow graph nodes until a barrier
construct is reached. All subtrees that are not translated to dataflow
graphs are kept as they are. The output of the translation pass is the
original AST where parallelizable regions have been replaced with
dataflow graphs and a call to \sys's runtime.

To identify possible opportunities for parallelism, the translation
pass extracts the parallelizability class of each command together
with its inputs and outputs. It achieves that by searching all
available PAL~\sx{ext} records for each command and resorts to
conservative defaults if a record for the command is not found. If the
command is \sta or parallelizable \pur, the translation pass initiates
a parallelizable region that is propagated up the tree.

%% Finally, whenever the algorithm encounters a command, it first
%% checks whether it is safe to parallelize it. A command is considered
%% safe for parallelization if (i) it belongs to the \sta or \pur distributability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this. % why couldn't we replicate this?
%% If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
%% If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.

%% A few illustrative cases of the analysis algorithm are shown in pseudocode below:

%% \begin{lstlisting}[language=python, float=h]
%%   def translate(node):
%%     ...
%%     elif(node.name == 'Pipe'):
%%       return pipe_graphs([translate(child)
%%         for child in node.children])
%%     elif(node.name == 'And'):
%%       node.children = [translate(child)
%%         for child in node.children]
%%       return node
%%     elif(node.name == 'Command'):
%%       if(not safe(node)):
%%         return make_ir_node(node, 'unsafe')
%%       else:
%%         return make_ir_node(node,
%%           find_category(node))
%%     ...
%% \end{lstlisting}

%% Whenever the algorithm encounters a distributable node, it recursively
%% translates its subcomponents.
%% It then merges them in a dataflow (sub-)graph, connecting the output of the first with the input of the second
%% % what first and second? we didn't say anything about first and second
%% Before the different dataflow graphs are connected, the
%% algorithm checks that at most one node in the graph writes at every
%% file, and that at most one node reads from a file that another node writes
%% to. This is important, in order to avoid inconsistent behaviour after
%% distributing nodes in the dataflow graph (\ie by introducing to
%% concurrent reads and writes to the same file when
%% parallelizing).

%% Whenever the algorithm encounters a barrier node,
%% it recursively translates its subcomponents, similarly to the distributable
%% case.
%% However, different from the distributable case, it does not merge them in a large dataflow graph.

%% Finally, whenever the algorithm encounters a command, it first
%% checks whether it is safe to parallelize it. A command is considered
%% safe for parallelization if (i) it belongs to the \sta or \pur distributability classes, and (ii) does not read from a read-once file (such as a pipe containing configuration data) as parallelization would break this. % why couldn't we replicate this?
%% If the command is safe, \sys creates a node containing that command  and adds it to a singleton dataflow graph.
%% If the command is unsafe, it is still added to the dataflow graph, but is not further parallelized by the optimizer.

An important consideration is that due to the highly dynamic nature of
the shell, there is information (such as the values of environment
variables) that are not known to \sys at the time of translation.
%% to the the safety properties that are mentioned above cannot be
%% soundly checked statically.
For safety purposes, \sys takes a conservative approach where it does
not parallelize nodes without complete information.  For example, the
translation pass cannot infer that an environment variable that is
given as an argument to a command is not a flag that changes its
parallelizability class,
%% two un-expand strings do not refer to the same file identifier,
and as a result does not parallelize this sub-expression.

\kk{Can we say anything more about the safety considerations? That (i)
  the translation pass ``checks'' that no nodes write/read or
  write/write to the same file (is that necessary?), that (ii) a
  command should not read from a read-once file. (is that necessary?)}

@@

%% \kk{Maybe guide all of this by showing an example of an output shell
%%   script?}

@@

\TODO{Should we mention that we empirically show that pash produces
  correct code?}
